Automated Lensing Learner - I: An automated Strong Lensing Identification Pipeline

### 2017-03-28 ms.tex
Changed LSST1 to LSST-best

Mock observations of lensing systems include the lens galaxy, lensed
images of the source galaxy, and galaxies along the line of sight.
Mock observations of non-lensing systems include all but the images of
a lensed source galaxy.  We convolve each of the $2\times 10,000$
train/test images and $2\times 1,000$ hold-out images to produce three
separate sets of mock ``observations''.  These each have equal numbers
of lens and non-lens containing systems: (1) HST-like observations,
(2) LSST-like observations over the span of one year, and (3)
LSST-like observations over the span of ten years.  We respectively
label these observations as HST, LSST1, and LSST10.

@@

To account for variations in seeing and sky-brightness over the course
of the survey, we draw each simulated exposure from a stochastic
distribution of these parameters. We then consider two different
strategies to use the simulated exposures. First, we build one
single-epoch image for each field (hereafter labeled as LSST1) by
keeping only the best seeing exposure.  Second, we build another
``worst-case'' stacked image by degrading all individual exposures to
match the one with the worst seeing and co-add all exposures to a
single image (hereafter labeled as LSST10).  These two sets of images
will allow us to investigate the trade-off between resolution and
signal-to-noise for our automated lens search.

@@

The middle column corresponds to the same simulated systems for
LSST10.  The right-most column corresponds to the simulated systems of
LSST1.  These images have resolution $45\times45~{\rm pixels}^2$ with
$0.18$~arcsec/pixel.  LSST10 images visually exhibit the improved
signal to noise ratio, recovering the arc feature, albeit at a much
lower resolution than with the HST-like image or the LSST1 image.  The
top images of LSST10 and LSST1 show a visible lensed source galaxy
image.  The bottom images are washed out in the bottom row, where the
magnification of the source galaxy is not as large.  The ground based
noise, PSF, and limited resolution of the LSST1 make visual giant arc
identification difficult, except in systems with the most magnified
source galaxies.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure*}[t]
\begin{center}
\includegraphics[width=2\columnwidth]{figures/tiles/compilations/telescope_compilation.pdf}
\caption{Left to right show example mock HST, LSST 10 year, and LSST 1
  year images.  The top row corresponds to a lensing system with a
  very visible arc signature, and the bottom row to a lensing system
  that is less obvious.  Example mock HST images have
  $n_\text{pix}\times n_\text{pix}=300\times300$.  Example mock LSST
  images have $n_\text{pix}\times n_\text{pix}=45\times45$.  The
  resolution and noise of a ground based telescope is noticeably
  worse. Visual identification of giant arcs in the LSST images in the
  bottom row is very difficult.}\label{fig:mockimages}
\end{center}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Furthermore, we investigate the influence of light from the lens
galaxies on the performance of our lens identification pipeline.  We
generate another set of each HST, LSST1, and LSST10 images without the
lens galaxy.  We respectively label these nHST, nLSST1, and nLSST10.

@@

Finally, the resolution of the overall image will also limit the
additional information that an increase in $N_\text{orient}$ will
provide.  From the grid search, the best case number of orientations
for each dataset is $N_\text{orient,HST}=9$,
$N_\text{orient,LSST10}=3$, and $N_\text{orient,LSST1}=5$.

@@

% Summary of ROC curves
Figure~\ref{fig:ROCcompilation} shows the ROC curves for models
trained using the entire 10,000 training sample, with best-case HOG
and regularization parameters.  The models have been evaluated on a
hold-out set of 1,000 images that were not used in the parameter
search.  We show the mock HST, LSST1, and LSST10 results respectively
in red, blue, and green.  Solid lines correspond to a model trained
and tested on images with the lensing galaxy.  Dashed lines correspond
to a model trained and tested on images where the lensing galaxy is
excluded from the mock observation, simulating an ideal modeling and
subtraction of the lensing galaxy, which has been one proposed method
to improve the identification of strong lensing systems.  The
corresponding AUC is listed in the legend.

@@

%  Our running over C - score and AUC as characterizations
Figure~\ref{fig:regularization} shows the model performance as a
function of regularization parameter for each data set HST, LSST, and
LSST10.  The solid and dotted blue lines respectively correspond to
the model performance on the test and training set, with the AUC as a
metric for performance.  In red, we show the train time as a function
of $C_{LogReg}$.

@@

% LSST10 and nLSST10 behavior
Figure~\ref{fig:trainsizeLSST} shows how the AUC depends on the log of
the size of the training set for both the LSST10 (LSST1) data in the
solid blue (red) line, and the nLSST10 (nLSST1) data in dashed blue
(red) line.  The AUC for models trained on the LSST10 data improves
almost linearly with the log of the training set size, increasing from
AUC=0.705 to AUC=0.788 when the train size is increased from
$2\times500$ lens/non-lens images to $2\times8000$ lens/non-lens
images.  However, for the nLSST10 data, where the lens galaxy has been
removed from the images, the improvement is less dramatic.  With the
same increase in training size, the AUC for nLSST10 changes from just
below 0.77 to just below 0.78.

@@

% LSST1 and nLSST1 behavior
In red, we see an analogous improvement for LSST1 and nLSST1.  Here,
the crossover happens at train size of $2x1000$, where an increase in
larger training data set size does not help improve the AUC for nLSST1.

@@

% Train time behavior for LSST1 and nLSST1
For LSST1, Figure~\ref{fig:traintimeLSST1} shows the same
relationships between train time and train size in blue, and the
optimal regulariztion parameter (or model complexity) in red.
Contrary to what we found when comparing LSST10 to nLSST10, nLSST1
requires more model complexity than LSST1.  Recall that LSST1 and
nLSST1 correspond to single epoch simulated exposures with the best
seeing; these images exhibit better resolution but worse signal to
noise.  The switch in required model complexity corresponds to a
trade-off between features from the lens galaxy providing additional
information or swamping the signal from a strong lensing arc.

% AUC vs. regularization on larger train/test set with comparable
% grid search in parameterization
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure}[t]
\begin{center}
\includegraphics[width=\columnwidth]{figures/trainsize/combined_LSST10_with_nLSST10.pdf}
\caption{Solid (dotted) blue line: AUC for models with varying size of
  the training set for LSST10 (nLSST10).  Solid (dotted) red line:
  Same for LSST1. The improvement of AUC scales roughly linearly with
  the log of the training set size.  However, LSST10 and LSST1 have a
  steeper improvement with data training size.  The performance of
  nLSST10 and nLSST1 models trained on smaller training data sets is
  better than the respective LSST10 and LSST1 models trained on the
  same size data, but the LSST10 and LSST1 models outperform with
  larger size training data set.}\label{fig:trainsizeLSST}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure}[t]
\begin{center}
  \includegraphics[width=\columnwidth]{figures/traintime/LSST10.pdf}}
\caption{Solid (dotted) blue line: Train time for models with varying
  size of the training set for LSST10 (nLSST10).  Train time roughly
  scales logarithmically with the train size, but the train time is
  also affected by model complexity.  Solid (dotted) red line: Best
  regularization parameter as a function of train size.  Note, LSST10
  requires more model complexity to exceed the performance of nLSST10
  (see blue solid and dotted lines in
  Figure~\ref{fig:trainsizeLSST}), and therefore requires more
  training time for continual increase in
  performance. }\label{fig:traintimeLSST10}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%            
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure}[t]
\begin{center}
  \includegraphics[width=\columnwidth]{figures/traintime/LSST1.pdf}}
\caption{Same as Figure~\ref{fig:traintimeLSST10}, but for LSST1.
  Here, nLSST1 requires more model complexity than LSST1.  Note, that
  LSST1 and nLSST1 have better seeing but reduced overall signal to
  noise than LSST10 and nLSST10.  This corresponds to better
  resolution, and therefore sharper edge features that prominently
  corespond to arcs.}\label{fig:traintimeLSST1}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In a forthcoming paper, we will discuss how class imbalance, or
differences between the lens-model parameter distributions in the
training and test sets affect model performance and will explore a
method to correct for this.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure*}[t]
  \centering 
  \mbox{
    \subfigure{\includegraphics[width=.66\columnwidth,trim=13 0 10 0,
        clip]{figures/lens_parameter_dependence/HST_Magnification_Velocity_Dispersion.pdf}}\hfill
    \subfigure{\includegraphics[width=.66\columnwidth,trim=13 0 10 0,
        clip]{figures/lens_parameter_dependence/LSST10_Magnification_Velocity_Dispersion.pdf}}\hfill
    \subfigure{\includegraphics[width=.66\columnwidth,trim=13 0 10 0,
        clip]{figures/lens_parameter_dependence/LSST1_Magnification_Velocity_Dispersion.pdf}\hfill}
  }\\ 
  \mbox{
    \subfigure{\includegraphics[width=.66\columnwidth,trim=13 0 10 0,
        clip]{figures/lens_parameter_dependence/HST_Magnification_z.pdf}}\hfill
    \subfigure{\includegraphics[width=.66\columnwidth,trim=13 0 10 0,
        clip]{figures/lens_parameter_dependence/LSST10_Magnification_z.pdf}}\hfill
    \subfigure{\includegraphics[width=.66\columnwidth,trim=13 0 10 0,
        clip]{figures/lens_parameter_dependence/LSST1_Magnification_z.pdf}\hfill}
  }\\ 
\caption{Left to right: Image classification score on HST, LSST10,
  LSST1 mock observed lensing images as a function of lensed image
  magnification.  Top row: Each panel is color coded by the lens
  velocity dispersion in km/s.  High velocity dispersion lens galaxies
  are likely to produce high magnification images of the source
  galaxies.  Magnification is the strongest indicator of how likely a
  lensing system will be successfully classified.  Bottom row: Here,
  we color code by the redshift of the lens galaxy.}\vspace{10pt}
\label{fig:scorevslensparams}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

% Realistic ratios
For images selected for a massive elliptical, the number of non-lenses
will outnumber lenses by at least 1000 to 1.  Therefore, a sample that
is $50\%$ pure requires a classifier with a false positive rate of
0.001.  Looking at the green line in Figure~\ref{fig:PRcompilation}, a
classification threshold for a pure sample only encapsulates
$\sim10\%$ (in recall) of all of our lenses before contamination from
non-lenses leaks into the selection.  A maximally large and pure
sample from our method would require further filtering, e.g. by
citizen science.

>>>>

Mock observations of lensing systems include the lens galaxy, lensed
images of the source galaxy, and galaxies along the line of sight.
Mock observations of non-lensing systems include all but the images of
a lensed source galaxy.  We convolve each of the $2\times 10,000$
train/test images and $2\times 1,000$ hold-out images to produce three
separate sets of mock ``observations''.  These each have equal numbers
of lens and non-lens containing systems: (1) HST-like observations,
(2) best single exposure LSST-like observations, and (3) LSST-like
observations over the span of ten years.  We respectively label these
observations as HST, LSST-best, and LSST10.

@@

To account for variations in seeing and sky-brightness over the course
of the survey, we draw each simulated exposure from a stochastic
distribution of these parameters. We then consider two different
strategies to use the simulated exposures. First, we build one
single-epoch image for each field (hereafter labeled as LSST-best) by
keeping only the best seeing exposure.  Second, we build another
``worst-case'' stacked image by degrading all individual exposures,
ten per filter per year, to match the one with the worst seeing and
co-add all exposures to a single image (hereafter labeled as LSST10).
These two sets of images will allow us to investigate the trade-off
between resolution and signal-to-noise for our automated lens search.

@@

The middle column corresponds to the same simulated systems for
LSST10.  The right-most column corresponds to the simulated systems of
LSST-best.  These images have resolution $45\times45~{\rm pixels}^2$ with
$0.18$~arcsec/pixel.  LSST10 images visually exhibit the improved
signal to noise ratio, recovering the arc feature, albeit at a much
lower resolution than with the HST-like image or the LSST-best image.
The top images of LSST10 and LSST-best show a visible lensed source
galaxy image.  The bottom images are washed out in the bottom row,
where the magnification of the source galaxy is not as large.  The
ground based noise, PSF, and limited resolution of the LSST-best make
visual giant arc identification difficult, except in systems with the
most magnified source galaxies.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure*}[t]
\begin{center}
\includegraphics[width=2\columnwidth]{figures/tiles/compilations/telescope_compilation.pdf}
\caption{Left to right show example mock HST, LSST 10 year, and
  LSST-best images.  The top row corresponds to a lensing system with a
  very visible arc signature, and the bottom row to a lensing system
  that is less obvious.  Example mock HST images have
  $n_\text{pix}\times n_\text{pix}=300\times300$.  Example mock LSST
  images have $n_\text{pix}\times n_\text{pix}=45\times45$.  The
  resolution and noise of a ground based telescope is noticeably
  worse. Visual identification of giant arcs in the LSST images in the
  bottom row is very difficult.}\label{fig:mockimages}
\end{center}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Furthermore, we investigate the influence of light from the lens
galaxies on the performance of our lens identification pipeline.  We
generate another set of each HST, LSST-best, and LSST10 images without the
lens galaxy.  We respectively label these nHST, nLSST-best, and nLSST10.

@@

Finally, the resolution of the overall image will also limit the
additional information that an increase in $N_\text{orient}$ will
provide.  From the grid search, the best case number of orientations
for each dataset is $N_\text{orient,HST}=9$,
$N_\text{orient,LSST10}=3$, and $N_\text{orient,LSST-best}=5$.

@@

% Summary of ROC curves
Figure~\ref{fig:ROCcompilation} shows the ROC curves for models
trained using the entire 10,000 training sample, with best-case HOG
and regularization parameters.  The models have been evaluated on a
hold-out set of 1,000 images that were not used in the parameter
search.  We show the mock HST, LSST-best, and LSST10 results respectively
in red, blue, and green.  Solid lines correspond to a model trained
and tested on images with the lensing galaxy.  Dashed lines correspond
to a model trained and tested on images where the lensing galaxy is
excluded from the mock observation, simulating an ideal modeling and
subtraction of the lensing galaxy, which has been one proposed method
to improve the identification of strong lensing systems.  The
corresponding AUC is listed in the legend.

@@

%  Our running over C - score and AUC as characterizations
Figure~\ref{fig:regularization} shows the model performance as a
function of regularization parameter for each data set HST, LSST-best, and
LSST10.  The solid and dotted blue lines respectively correspond to
the model performance on the test and training set, with the AUC as a
metric for performance.  In red, we show the train time as a function
of $C_{LogReg}$.

@@

% LSST10 and nLSST10 behavior
Figure~\ref{fig:trainsizeLSST} shows how the AUC depends on the log of
the size of the training set for both the LSST10 (LSST-best) data in the
solid blue (red) line, and the nLSST10 (nLSST-best) data in dashed blue
(red) line.  The AUC for models trained on the LSST10 data improves
almost linearly with the log of the training set size, increasing from
AUC=0.705 to AUC=0.788 when the train size is increased from
$2\times500$ lens/non-lens images to $2\times8000$ lens/non-lens
images.  However, for the nLSST10 data, where the lens galaxy has been
removed from the images, the improvement is less dramatic.  With the
same increase in training size, the AUC for nLSST10 changes from just
below 0.77 to just below 0.78.

@@

% LSST1 and nLSST1 behavior
In red, we see an analogous improvement for LSST-best and nLSST-best.
Here, the crossover happens at train size of $2x1000$, where an
increase in larger training data set size does not help improve the
AUC for nLSST-best.

@@

% Train time behavior for LSST1 and nLSST1
For LSST-best, Figure~\ref{fig:traintimeLSST1} shows the same
relationships between train time and train size in blue, and the
optimal regulariztion parameter (or model complexity) in red.
Contrary to what we found when comparing LSST10 to nLSST10, nLSST-best
requires more model complexity than LSST-best.  Recall that LSST-best and
nLSST-best correspond to single epoch simulated exposures with the best
seeing; these images exhibit better resolution but worse signal to
noise.  The switch in required model complexity corresponds to a
trade-off between features from the lens galaxy providing additional
information or swamping the signal from a strong lensing arc.

% AUC vs. regularization on larger train/test set with comparable
% grid search in parameterization
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure}[t]
\begin{center}
\includegraphics[width=\columnwidth]{figures/trainsize/combined_LSST10_with_nLSST10.pdf}
\caption{Solid (dotted) blue line: AUC for models with varying size of
  the training set for LSST10 (nLSST10).  Solid (dotted) red line:
  Same for LSST-best. The improvement of AUC scales roughly linearly with
  the log of the training set size.  However, LSST10 and LSST-best have a
  steeper improvement with data training size.  The performance of
  nLSST10 and nLSST-best models trained on smaller training data sets is
  better than the respective LSST10 and LSST-best models trained on the
  same size data, but the LSST10 and LSST-best models outperform with
  larger size training data set.}\label{fig:trainsizeLSST}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure}[t]
\begin{center}
  \includegraphics[width=\columnwidth]{figures/traintime/LSST10.pdf}}
\caption{Solid (dotted) blue line: Train time for models with varying
  size of the training set for LSST10 (nLSST10).  Train time roughly
  scales logarithmically with the train size, but the train time is
  also affected by model complexity.  Solid (dotted) red line: Best
  regularization parameter as a function of train size.  Note, LSST10
  requires more model complexity to exceed the performance of nLSST10
  (see blue solid and dotted lines in
  Figure~\ref{fig:trainsizeLSST}), and therefore requires more
  training time for continual increase in
  performance. }\label{fig:traintimeLSST10}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%            
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure}[t]
\begin{center}
  \includegraphics[width=\columnwidth]{figures/traintime/LSST1.pdf}}
\caption{Same as Figure~\ref{fig:traintimeLSST10}, but for LSST-best.
  Here, nLSST-best requires more model complexity than LSST-best.  Note, that
  LSST-best and nLSST-best have better seeing but reduced overall signal to
  noise than LSST10 and nLSST10.  This corresponds to better
  resolution, and therefore sharper edge features that prominently
  corespond to arcs.}\label{fig:traintimeLSST1}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In a forthcoming paper, we will discuss how class imbalance, or
differences between the lens-model parameter distributions in the
training and test sets affect model performance and will explore a
method to correct for this.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure*}[t]
  \centering 
  \mbox{
    \subfigure{\includegraphics[width=.66\columnwidth,trim=13 0 10 0,
        clip]{figures/lens_parameter_dependence/HST_Magnification_Velocity_Dispersion.pdf}}\hfill
    \subfigure{\includegraphics[width=.66\columnwidth,trim=13 0 10 0,
        clip]{figures/lens_parameter_dependence/LSST10_Magnification_Velocity_Dispersion.pdf}}\hfill
    \subfigure{\includegraphics[width=.66\columnwidth,trim=13 0 10 0,
        clip]{figures/lens_parameter_dependence/LSST1_Magnification_Velocity_Dispersion.pdf}\hfill}
  }\\ 
  \mbox{
    \subfigure{\includegraphics[width=.66\columnwidth,trim=13 0 10 0,
        clip]{figures/lens_parameter_dependence/HST_Magnification_z.pdf}}\hfill
    \subfigure{\includegraphics[width=.66\columnwidth,trim=13 0 10 0,
        clip]{figures/lens_parameter_dependence/LSST10_Magnification_z.pdf}}\hfill
    \subfigure{\includegraphics[width=.66\columnwidth,trim=13 0 10 0,
        clip]{figures/lens_parameter_dependence/LSST1_Magnification_z.pdf}\hfill}
  }\\ 
\caption{Left to right: Image classification score on HST, LSST10,
  LSST-best mock observed lensing images as a function of lensed image
  magnification.  Top row: Each panel is color coded by the lens
  velocity dispersion in km/s.  High velocity dispersion lens galaxies
  are likely to produce high magnification images of the source
  galaxies.  Magnification is the strongest indicator of how likely a
  lensing system will be successfully classified.  Bottom row: Here,
  we color code by the redshift of the lens galaxy.}\vspace{10pt}
\label{fig:scorevslensparams}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

% Realistic ratios
For images selected for a massive elliptical, the number of non-lenses
will outnumber lenses by at least 1000 to 1.  Therefore, a sample that
is $50\%$ pure requires a classifier with a false positive rate of
0.001.  Looking at the green line in Figure~\ref{fig:PRcompilation}, a
classification threshold for a pure sample only encapsulates
$\sim10\%$ (in recall) of all of our lenses before contamination from
non-lenses leaks into the selection.  While information from other
bands will certainly improve the model performance, a maximally large
and pure sample from our method would likely require further
filtering, e.g. by citizen science.
